Secure Composition of Robust and Optimising Compilers

### 2023-06-20 popl/main.tex
Compiler

\begin{center}
  \judgbox{\pstepto{\runtimetermvar}{\runtimetermvar[']}{\event}}{,,Primitive step from runtime--term $\runtimetermvar$ to $\runtimetermvar[']$ emitting event $\event$.''}
  \typerule{$e-\text{get}-\in$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\in\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lget{x}{n}}}{\rtt{\statevar}{\heap[^{\sandboxtag}](\loc+n)}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-in}
  %
  \typerule{$e-\text{get}-\notin$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\notin\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lget{x}{n}}}{\rtt{\statevar}{1729}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-notin}
  %
  \typerule{$e-\text{set}-\in$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\in\statevar.\heap[^{\sandboxtag}] &
    \heap[^{\sandboxtag}_1] = \seth{\statevar.\heap[^{\sandboxtag}]}{\loc+n}{\valueexpr}
  }{
    \pstepto{\rtt{\statevar}{\lset{x}{n}{\valueexpr}}}{\rtt{\statevar\subst{\heap[^{\sandboxtag}_1]}{\heap[^{\sandboxtag}]}}{\valueexpr}}{\tup{\ev{Set}\ \loc\ n\ \valueexpr}{\sandboxtag}}
  }{e-set-in}
  %
  \typerule{$e-\text{set}-\notin$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\notin\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lset{x}{n}{\valueexpr}}}{\rtt{\statevar}{\valueexpr}}{\tup{\ev{Set}\ \loc\ n\ \valueexpr}{\sandboxtag}}
  }{e-set-notin}
  %
  \typerule{$e-\text{new}$}{
    \fresh{\statevar}{z} &
    \fresh{\statevar}{\loc} &
    \heap[^{\sandboxtag}_1]=\growh{\statevar.\heap[^{\sandboxtag}]}{n} &
    \store[_1]=\storeel{z}{\loc}{\statevar.\sandboxtag}{\poisonless}{n},\statevar.\store
  }{
    \pstepto{\rtt{\statevar}{\lnew{x}{n}{\expr}}}{\rtt{\statevar\subst{\heap[^{\sandboxtag}]}{\heap[^{\sandboxtag}_1]}\subst{\store}{\store[_1]}}{\expr\subst{z}{x}}}{\tup{\ev{Alloc}\ \loc\ n}{\sandboxtag}}
  }{e-new}
  %
  \typerule{$e-\text{dealloc}$}{
    \statevar.\store(x) = \tup{\loc}{\statevar.\sandboxtag,\poison,n} &
    \store[_1]=\seth{\statevar.\store}{x}{\tup{\loc}{\statevar.\sandboxtag,\poisoned,n}}
  }{
    \pstepto{\rtt{\statevar}{\ldelete{x}}}{\rtt{\statevar\subst{\store}{\store[_1]}}{0}}{\tup{\ev{Dealoc}\ \loc}{\sandboxtag}}
  }{e-dealloc}
  %
  \typerule{$e-\text{abort}$}{
  }{
    \pstepto{\rtt{\statevar}{\labort}}{\lightning}{\lightning}
  }{e-abort}
\end{center}

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}&\bnfdef\src{\natt} \mid \src{\ptrn} \hspace{0.5cm}
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Type~Environment)}~\src{\typenv}&\bnfdef\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \hspace{0.5cm}
    \mi{(Types)}~\src{\type}\bnfdef\src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]}
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

\subsection{$\trg{L}$: An Unsafe Language}\label{subsec:ltms}

@@

\MKin{compiler $\src{L}\to\trg{L}$}
\MKin{x-lang relations}
\MKin{backtranslation}
\MKin{rtp of tms}

@@

\subsection{$\obj{L_{\scctsafe}}$: A Language with Access to a Model Specific Register}\label{subsec:ltms}

>>>>

\begin{center}
  \judgbox{\pstepto{\runtimetermvar}{\runtimetermvar[']}{\event}}{,,Primitive step from runtime--term $\runtimetermvar$ to $\runtimetermvar[']$ emitting event $\event$.''}
  \typerule{$e-\text{get}-\in$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\in\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lget{x}{n}}}{\rtt{\statevar}{\heap[^{\sandboxtag}](\loc+n)}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-in}
  %
  \typerule{$e-\text{get}-\notin$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\notin\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lget{x}{n}}}{\rtt{\statevar}{1729}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-notin}
  %
  \typerule{$e-\text{set}-\in$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\in\statevar.\heap[^{\sandboxtag}] &
    \heap[^{\sandboxtag}_1] = \seth{\statevar.\heap[^{\sandboxtag}]}{\loc+n}{\valueexpr}
  }{
    \pstepto{\rtt{\statevar}{\lset{x}{n}{\valueexpr}}}{\rtt{\statevar\subst{\heap[^{\sandboxtag}_1]}{\heap[^{\sandboxtag}]}}{\valueexpr}}{\tup{\ev{Set}\ \loc\ n\ \valueexpr}{\sandboxtag}}
  }{e-set-in}
  %
  \typerule{$e-\text{set}-\notin$}{
    \sandboxtag=\statevar.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} \\
    \loc+n\notin\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\statevar}{\lset{x}{n}{\valueexpr}}}{\rtt{\statevar}{\valueexpr}}{\tup{\ev{Set}\ \loc\ n\ \valueexpr}{\sandboxtag}}
  }{e-set-notin}
  %
  \typerule{$e-\text{new}$}{
    \fresh{\statevar}{z} &
    \fresh{\statevar}{\loc} &
    \heap[^{\sandboxtag}_1]=\growh{\statevar.\heap[^{\sandboxtag}]}{n} &
    \store[_1]=\storeel{z}{\loc}{\statevar.\sandboxtag}{\poisonless}{n},\statevar.\store
  }{
    \pstepto{\rtt{\statevar}{\lnew{x}{n}{\expr}}}{\rtt{\statevar\subst{\heap[^{\sandboxtag}]}{\heap[^{\sandboxtag}_1]}\subst{\store}{\store[_1]}}{\expr\subst{z}{x}}}{\tup{\ev{Alloc}\ \loc\ n}{\sandboxtag}}
  }{e-new}
  %
  \typerule{$e-\text{dealloc}$}{
    \statevar.\store(x) = \tup{\loc}{\statevar.\sandboxtag,\poison,n} &
    \store[_1]=\seth{\statevar.\store}{x}{\tup{\loc}{\statevar.\sandboxtag,\poisoned,n}}
  }{
    \pstepto{\rtt{\statevar}{\ldelete{x}}}{\rtt{\statevar\subst{\store}{\store[_1]}}{0}}{\tup{\ev{Dealloc}\ \loc}{\sandboxtag}}
  }{e-dealloc}
  %
  \typerule{$e-\text{abort}$}{
  }{
    \pstepto{\rtt{\statevar}{\labort}}{\lightning}{\lightning}
  }{e-abort}
\end{center}

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}&\bnfdef\src{\natt} \mid \src{\ptrn} \hspace{0.5cm}
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Type~Environment)}~\src{\typenv}&\bnfdef\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \hspace{0.5cm}
    \mi{(Types)}~\src{\type}\bnfdef\src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]} \\
    \mi{(Symbols)}~\src{\function}&\bnfdef\src{\lfunction{g}{x:\type[_{e}]\to\type[_{e}]}{\expr}} \\
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

\subsection{$\trg{L}$: An Unsafe Language}\label{subsec:lsms}

\begin{gather*}
  \begin{aligned}
    \mi{(Expressions)}~\trg{\expr}\bnfdef\ \dots \mid \trg{\lispoisoned{x}} \mid \trg{\lpair{\expr[_{1}]}{\expr[_{2}]}} &\mid \trg{\lproja{\expr}} \mid \trg{\lprojb{\expr}} \mid \trg{\lhast{\expr}{\type}} \hspace{0.5cm}
    \mi{(Types)}~\trg{\type}\bnfdef\ \trg{\natt} \mid \trg{\natt\times\natt} \\
    \mi{(Values)}~\trg{\valueexpr}&\bnfdef\ \dots \mid \trg{\lpair{n_{1}}{n_{2}}}
  \end{aligned}
\end{gather*}
\begin{center}
  \typerule{e-$\trg{\lispoisoned{}}$-yes}{
    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisoned,n}}
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-ispoisoned-yes}
  %
  \typerule{e-$\trg{\lispoisoned{}}$-no}{
    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisonless,n}}
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-ispoisoned-no}
  %
  \typerule{e-proj1}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lproja{\lpair{n_1}{n_2}}}}}{\rtt{\trg{\statevar}}{\trg{n_1}}}{\trg{\emptyevent}}
  }{e-proj1}
  %
  \typerule{e-$\trg{\lhast{n}{\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-n-has-nat}
  %
  \typerule{e-$\trg{\lhast{x}{\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{x}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-x-has-nat}
  %
  \typerule{e-$\trg{\lhast{n}{\natt\times\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt\times\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-n-has-nattimesnat}
\end{center}

The $\trg{L}$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes,tr:e-ispoisoned-no}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-x-has-nat,tr:e-n-has-nattimesnat}).
As before, not all rules are shown due to their straightforwardness.
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.

@@

\begin{center}
  $$
  \begin{array}{rcl}
    \cca(\src{\lbinop{\expr[_{1}]}{\expr[_{2}]}}) & = & \lbinop[\trg]{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr[_{2}]})\right]} \\ [0.33cm]
    \cca(\src{\lfunction{g}{x:\natt\to\type[_{e}]}{\expr}}) & = & \lfunction[\trg]{\trg{g}}{\trg{x}}{\lifz[\trg]{\trg{\lhast{x}{\natt}}}{%
                                                                                                \left[\cca(\src{\expr})\right] %
                                                                                                 }{\labort[\trg]}}
  \end{array}
  $$
\end{center}

The only case where the compiler does something interesting is for top-level functions.
Here, a dynamic typecheck is inserted to respect the calling-convention.
Since $\trg{L}$ has no static typechecks, it could happen that a bogus context $\trg{\library_{\ctx}}$ calls this function expecting a natural number with $\trg{\lpair{17}{29}}$.
The compiler ensures that execution does not proceed in cases like that.

\MKin{x-lang relations}
\MKin{backtranslation}
\MKin{rtp of tms}

@@

\subsection{$\obj{L_{\scctsafe}}$: A Language with Access to a Model Specific Register}\label{subsec:lscct}
